---
title: "Walmart Sales Prediction"
author: "Ariel Li, Jiaying Du, Sylvie Pan"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load the packages
suppressMessages({
  library(data.table)
  library(dplyr)
  library(tidyverse)
  library(psych)
  library(scales)
  library(lubridate)
  library(ggplot2) 
  library(hrbrthemes)
  library(plotly)
  library(forecast) 
  library(prophet) 
})

set.seed(42)
```

### Data

calendar.csv - Contains information about the dates on which the products are sold.
sell_prices.csv - Contains information about the price of the products sold per store and date.
sales_train_validation.csv - Contains the historical daily unit sales data per product and store [d_1 - d_1913].

```{r}
# read the data
calendar <- fread('data/calendar.csv')
price <- fread('data/sell_prices.csv')
sales <- fread('data/sales_train_validation.csv')
```

```{r}
head(calendar)
head(price)
```

```{r}
# train test split
basic <- select(sales, 1:6)
date <- select(sales, -1:-6)
date_num <- dim(date)[2]
test_size <- 28
train_size <- date_num - 28
train <- select(date, 1:all_of(train_size))
test <- select(date, (all_of(train_size)+1):all_of(date_num))
train <- cbind(basic, train)
test <- cbind(basic, test)
```

The data range from 2011-01-29 to 2016-06-19.
```{r}
calendar$date <- as.Date(calendar$date, format='%Y-%m-%d')
calendar %>% summarize(min_date=min(date), max_date=max(date))
```

The data range for the training data set is from 2011-01-29 to 2016-03-27, and the data range for the test set is from 2016-03-28 to 2016-04-24.
```{r}
train_date <- calendar$date[1:train_size]
test_date <- calendar$date[(train_size+1):(train_size+28)]
min(train_date)
max(train_date)
min(test_date)
max(test_date)
```

In our dataset, we have 3 categories of 3049 items in 7 departments. Those items were sold in 10 stores of 3 states.
```{r}
# count unique item, dept, category, store, and state
length(unique(train[['item_id']]))
length(unique(train[['dept_id']]))
length(unique(train[['cat_id']]))
length(unique(train[['store_id']]))
length(unique(train[['state_id']]))
```

```{r}
(top_item <- head(count(train, item_id, sort = TRUE), 5))
(top_dept <- count(train, dept_id, sort = TRUE))
(top_cat <- count(train, cat_id, sort = TRUE))
(top_store <- count(train, store_id, sort = TRUE))
(top_state <- count(train, state_id, sort = TRUE))
```

FOODS category and FOODS_3 department have the largest number of products.
```{r}
ggplot(train, aes(x=dept_id)) +
  geom_bar(fill='steelblue') + 
  ggtitle('Item Count by Department') +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(train, aes(x=cat_id)) +
  geom_bar(fill='steelblue') + 
  ggtitle('Item Count by Category') +
  theme(plot.title = element_text(hjust = 0.5))
```

There are 4 stores in CA, 3 in TX, and 3 in WI.
```{r}
(store_state <- train %>% group_by(state_id) %>% summarize(unique_stores=n_distinct(store_id)))

ggplot(store_state, aes(state_id, unique_stores)) +
  geom_col(fill='steelblue') + 
  ggtitle('Number of Stores by State') +
  theme(plot.title = element_text(hjust = 0.5))
```

<!-- # ```{r} -->
<!-- # # melt sales -->
<!-- # train <- melt(train, measure.vars=patterns('^d_'), variable.name='d', value.name='sales') -->
<!-- # test <- melt(test, measure.vars=patterns('^d_'), variable.name='d', value.name='sales') -->
<!-- # head(train) -->
<!-- # head(test) -->
<!-- # ``` -->

<!-- # ```{r} -->
<!-- # # join tables -->
<!-- # train %>%  -->
<!-- #   left_join(calendar, by = 'd') %>%  -->
<!-- #   left_join(price, by = c('store_id', 'item_id', 'wm_yr_wk')) -->

<!-- # test %>%  -->
<!-- #   left_join(calendar, by = 'd') %>%  -->
<!-- #   left_join(price, by = c('store_id', 'item_id', 'wm_yr_wk')) -->
<!-- # ``` -->

<!-- # ```{r} -->
<!-- # write.csv(train, 'train.csv') -->
<!-- # write.csv(test, 'test.csv') -->
<!-- # ``` -->

```{r}
# read the data
train <- fread('data/train.csv')
test <- fread('data/test.csv')
```

```{r}
head(train)
head(test)
```

## Sales

### General Sales

We can see an upward trend from 2011 to 2016. We can also observe seasonality with annual peaks around September and October, and a dip on Christmas every year when all stores closed. 

```{r}
daily_sales_df <- train %>% group_by(date) %>% summarize(daily_sales=sum(sales))
daily_sales_df$date <- as.Date(daily_sales_df$date)
head(daily_sales_df)
```

```{r}
fig <- daily_sales_df %>%
  ggplot(aes(x=date, y=daily_sales, group=1)) +
  geom_area(fill='#69b3a2', alpha=0.5) +
  geom_line(color='#69b3a2') +
  geom_smooth(method='lm', formula=y~x, se=FALSE, size=0.5, color='gray20') +
  labs(x='Date', y='Sales', title='Daily Sales') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```

### Sales by Item

There are 3049 distinct items in the dataset. We will randomly selected 20 items to plot here.
```{r}
set.seed(42)
random_items <- sample(unique(train$item_id), 20)

item_sales_df <- train %>% 
  filter(item_id %in% random_items) %>%
  group_by(date, item_id) %>% 
  summarize(item_sales=sum(sales))

item_sales_df$date <- as.Date(item_sales_df$date)

fig <- item_sales_df %>%
  ggplot(aes(x=date, y=item_sales, col=item_id)) +
  geom_line() +
  labs(x='Date', y='Sales', title='Daily Sales by Item') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Sales by Department
```{r}
dept_sales_df <- train %>% group_by(date, dept_id) %>% summarize(dept_sales=sum(sales))
dept_sales_df$date <- as.Date(dept_sales_df$date)

fig <- dept_sales_df %>%
  ggplot(aes(x=date, y=dept_sales, col=dept_id)) +
  geom_line() +
  labs(x='Date', y='Sales', title='Daily Sales by Department') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Sales by Category
```{r}
cat_sales_df <- train %>% group_by(date, cat_id) %>% summarize(cat_sales=sum(sales))
cat_sales_df$date <- as.Date(cat_sales_df$date)

fig <- cat_sales_df %>%
  ggplot(aes(x=date, y=cat_sales, col=cat_id)) +
  geom_line() +
  labs(x='Date', y='Sales', title='Daily Sales by Product Category') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Sales by Store
```{r}
store_sales_df <- train %>% group_by(date, store_id) %>% summarize(store_sales=sum(sales))
store_sales_df$date <- as.Date(store_sales_df$date)

fig <- store_sales_df %>%
  ggplot(aes(x=date, y=store_sales, col=store_id)) +
  geom_line() +
  labs(x='Date', y='Sales', title='Daily Sales by Store') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Sales by State
```{r}
state_sales_df <- train %>% group_by(date, state_id) %>% summarize(state_sales=sum(sales))
state_sales_df$date <- as.Date(state_sales_df$date)

fig <- state_sales_df %>%
  ggplot(aes(x=date, y=state_sales, col=state_id)) +
  geom_line() +
  labs(x='Date', y='Sales', title='Daily Sales by State') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


## Price

### Price by Item

There are 3049 distinct items in the dataset. We will randomly selected 20 items to plot here.
```{r}
set.seed(42)
random_items <- sample(unique(train$item_id), 20)

item_price_df <- train %>% 
  filter(item_id %in% random_items) %>%
  group_by(date, item_id) %>% 
  summarize(item_price=mean(sell_price))

item_price_df$date <- as.Date(item_price_df$date)

fig <- item_price_df %>%
  ggplot(aes(x=date, y=item_price, col=item_id)) +
  geom_line() +
  labs(x='Date', y='Price', title='Item Daily Average Price') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Price by Department
```{r}
dept_price_df <- train %>% group_by(date, dept_id) %>% summarize(dept_price=mean(sell_price, 
na.rm=TRUE))

dept_price_df$date <- as.Date(dept_price_df$date)

fig <- dept_price_df %>%
  ggplot(aes(x=date, y=dept_price, col=dept_id)) +
  geom_line() +
  labs(x='Date', y='Price', title='Department Daily Average Price') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Price by Category
```{r}
cat_price_df <- train %>% group_by(date, cat_id) %>% summarize(cat_price=mean(sell_price, 
na.rm=TRUE))

cat_price_df$date <- as.Date(cat_price_df$date)

fig <- cat_price_df %>%
  ggplot(aes(x=date, y=cat_price, col=cat_id)) +
  geom_line() +
  labs(x='Date', y='Price', title='Daily Average Price by Category') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Price by Store
```{r}
store_price_df <- train %>% group_by(date, store_id) %>% summarize(store_price=mean(sell_price, 
na.rm=TRUE))

store_price_df$date <- as.Date(store_price_df$date)

fig <- store_price_df %>%
  ggplot(aes(x=date, y=store_price, col=store_id)) +
  geom_line() +
  labs(x='Date', y='Price', title='Daily Average Price by Store') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```


### Price by State
```{r}
state_price_df <- train %>% group_by(date, state_id) %>% summarize(state_price=mean(sell_price, 
na.rm=TRUE))

state_price_df$date <- as.Date(state_price_df$date)

fig <- state_price_df %>%
  ggplot(aes(x=date, y=state_price, col=state_id)) +
  geom_line() +
  labs(x='Date', y='Price', title='Daily Average Price by State') +
  scale_x_date(date_breaks='3 month', date_labels='%b %y') +
  theme_ipsum() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

ggplotly(fig, dynamicTicks=TRUE)
```




















