---
title: "Walmart Sales Prediction"
author: "Ariel Li, Jiaying Du, Sylvie Pan"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load the packages
suppressMessages({
  library(data.table)
  library(dplyr)
  library(tidyverse)
  library(psych)
  library(scales)
  library(lubridate)
  library(ggplot2) 
  library(forecast) 
  library(prophet) 
})

set.seed(42)
```

calendar.csv - Contains information about the dates on which the products are sold.
sell_prices.csv - Contains information about the price of the products sold per store and date.
sales_train_validation.csv - Contains the historical daily unit sales data per product and store [d_1 - d_1913].

```{r}
# read the data
calendar <- fread('data/calendar.csv')
price <- fread('data/sell_prices.csv')
sales <- fread('data/sales_train_validation.csv')
```

```{r}
head(calendar)
head(price)
```

```{r}
# train test split
basic <- select(sales, 1:6)
date <- select(sales, -1:-6)
date_num <- dim(date)[2]
test_size <- 28
train_size <- date_num - 28
train <- select(date, 1:all_of(train_size))
test <- select(date, (all_of(train_size)+1):all_of(date_num))
train <- cbind(basic, train)
test <- cbind(basic, test)
```

The data range from 2011-01-29 to 2016-06-19.
```{r}
calendar$date <- as.Date(calendar$date, format='%Y-%m-%d')
calendar %>% summarize(min_date=min(date), max_date=max(date))
```

The data range for the training data set is from 2011-01-29 to 2016-03-27, and the data range for the test set is from 2016-03-28 to 2016-04-24.
```{r}
train_date <- calendar$date[1:train_size]
test_date <- calendar$date[(train_size+1):(train_size+28)]
min(train_date)
max(train_date)
min(test_date)
max(test_date)
```

In our dataset, we have 3 categories of 3049 items in 7 departments. Those items were sold in 10 stores of 3 states.
```{r}
# count unique item, dept, category, store, and state
length(unique(train[['item_id']]))
length(unique(train[['dept_id']]))
length(unique(train[['cat_id']]))
length(unique(train[['store_id']]))
length(unique(train[['state_id']]))
```

```{r}
(top_item <- head(count(train, item_id, sort = TRUE), 5))
(top_dept <- count(train, dept_id, sort = TRUE))
(top_cat <- count(train, cat_id, sort = TRUE))
(top_store <- count(train, store_id, sort = TRUE))
(top_state <- count(train, state_id, sort = TRUE))
```

FOODS category and FOODS_3 department have the largest number of products.
```{r}
ggplot(train, aes(x=dept_id)) +
  geom_bar(fill='steelblue') + 
  ggtitle('Item Count by Department') +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(train, aes(x=cat_id)) +
  geom_bar(fill='steelblue') + 
  ggtitle('Item Count by Category') +
  theme(plot.title = element_text(hjust = 0.5))
```

There are 4 stores in CA, 3 in TX, and 3 in WI.
```{r}
(store_state <- train %>% group_by(state_id) %>% summarize(unique_stores=n_distinct(store_id)))

ggplot(store_state, aes(state_id, unique_stores)) +
  geom_col(fill='steelblue') + 
  ggtitle('Number of Stores by State') +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# melt sales
train <- melt(train, measure.vars=patterns('^d_'), variable.name='d', value.name='sales')
test <- melt(test, measure.vars=patterns('^d_'), variable.name='d', value.name='sales')
head(train)
head(test)
```

```{r}
# join tables
train %>% 
  left_join(calendar, by = 'd') %>% 
  left_join(price, by = c('store_id', 'item_id', 'wm_yr_wk'))

test %>% 
  left_join(calendar, by = 'd') %>% 
  left_join(price, by = c('store_id', 'item_id', 'wm_yr_wk'))
```

```{r}
write.csv(train, 'train.csv')
write.csv(test, 'test.csv')
```




